% !TEX root = thesis.tex

\chapter{Continuous Program Optimization Techniques}
\label{ch:continuous}

In this chapter, we focus on a fundamental aspect for deploying adaptive optimization techniques in runtime systems: the {\em On-Stack Replacement} (OSR) problem. OSR consists in dynamically transferring execution between different versions of a function at run time. Modern virtual machines implement OSR to {\em continuously} optimize a program as it executes, for instance by interrupting a long-running function and recompiling it at a higher optimization level, or by replacing a function version with another when a speculative assumption made during its compilation no longer holds.

In the first part of the chapter, we present a platform-independent framework for OSR that introduces novel ideas and combines features of existing techniques that no previous solution provided simultaneously. In particular, we introduce a {\em compensation code} abstraction that increases the flexibility of OSR mechanisms, and we present an OSR implementation for the LLVM MCJIT compiler that allows OSR to happen at arbitrary locations in a function.

In the second part, we make a first step towards a provably sound methodological framework for OSR. We formalize the concept of {\em multi-version program}, and we identify sufficient conditions for the correctness of an OSR transition. We also devise an algorithm for automatically generating compensation code possibly required to realign program state in the presence of several common compiler optimizations.

\input{osrkit}
\input{osr-correctness}

\section{Summary}
On-stack replacement is a staple of modern adaptive optimization systems, and has tended to be restricted to a few of the most advanced language runtimes that have seen enormous commercial investments. We present an OSR framework that combines some of the best practices in the literature with novel features, and that should allow developers to leverage OSR in their runtimes without the daunting prospect of building the mechanisms from the ground up. We introduce a theoretical framework that might allow the community to make progress in reasoning about the underpinnings of OSR and its soundness in the face of important code transformations. We evaluate implementations of our ideas in LLVM in \mysection\ref{se:eval-osrkit,se:eval-OSR-alC}.